---
title: "Modeling continuous cell-level covariates"
subtitle: 'Collapse using mean value for pseudobulk data'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
documentclass: article
vignette: >
  %\VignetteIndexEntry{Modeling continuous cell-level covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
output:
  html_document:
    toc: true
    toc_float: true
---

<!---

cd /Users/gabrielhoffman/workspace/repos/dreamlet/vignettes

rmarkdown::render("cell_level_covariates.Rmd")


--->


<style>
body {
text-align: justify}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png"),
  cache = TRUE)
```

Since read counts are summed across cells in a pseudobulk approach, modeling continuous cell-level covariates also requires a collapsing step.  Here we summarize the values of a variable from a set of cells using the mean, and store the value for each cell type.  Including these variables in a regression formula uses the summarized values from the corresponding cell type.   

We demonstrate this feature on a lightly modified analysis of PBMCs from 8 individuals stimulated with interferon-β [Kang, et al, 2018, Nature Biotech](https://www.nature.com/articles/nbt.4042).


# Standard processing

Here is the code from the main vignette:

```{r preprocess.data}
library(dreamlet)
library(muscat)
library(ExperimentHub)
library(scater)

# Download data, specifying EH2259 for the Kang, et al study
eh <- ExperimentHub()
sce <- eh[["EH2259"]]

# only keep singlet cells with sufficient reads
sce <- sce[rowSums(counts(sce) > 0) > 0, ]
sce <- sce[,colData(sce)$multiplets == 'singlet']

# compute QC metrics
qc <- perCellQCMetrics(sce)

# remove cells with few or many detected genes
ol <- isOutlier(metric = qc$detected, nmads = 2, log = TRUE)
sce <- sce[, !ol]

# set variable indicating stimulated (stim) or control (ctrl)
sce$StimStatus = sce$stim
```

In many datasets, continuous cell-level variables could be mapped reads, gene count, mitochondrial rate, etc. There are no continuous cell-level variables in this dataset, so we can simulate two from a normal distribution:
```{r variable}
sce$value1 = rnorm(ncol(sce))
sce$value2 = rnorm(ncol(sce))
```

# Pseudobulk
Now compute the pseudobulk using standard code:

```{r aggregate}
sce$id <- paste0(sce$StimStatus, sce$ind)

# Create pseudobulk 
pb <- aggregateToPseudoBulk(sce,
    assay = "counts",     
    cluster_id = "cell",  
    sample_id = "id",
    verbose = FALSE)
```

The means per variable, cell type, and sample are stored in the pseudobulk `SingleCellExperiment` object:

```{r aggr_means}
metadata(pb)$aggr_means
```

# Analysis
Including these variables in a regression formula uses the summarized values from the corresponding cell type.  This happens behind the scenes, so the user doesn't need to distinguish bewteen sample-level variables stored in `colData(pb)` and cell-level variables stored in `metadata(pb)$aggr_means`.

Variance partition and hypothesis testing proceeds as ususal:

```{r processAssays, fig.height=12}
form = ~ StimStatus + value1 + value2

# Normalize and apply voom/voomWithDreamWeights
res.proc = processAssays( pb, form, min.count=5)

# run variance partitioning analysis
vp.lst = fitVarPart( res.proc, form)

# Summarize variance fractions genome-wide for each cell type
plotVarPart(vp.lst, label.angle=60) 

# Differential expression analysis within each assay
res.dl = dreamlet( res.proc, form)

# dreamlet results include coefficients for value1 and value2
res.dl
```




# Session Info
<details>
```{r session, echo=FALSE}
sessionInfo()
```
</details>






